Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-25 Paper/Paper.tex
slight improvement
Dan stuff added

% Three addresses
% --------------
\threeauthors
  {First author} {Affiliation1 \\ {\tt author1@ismir.edu}}
  {Second author} {\bf Retain these fake authors in\\\bf submission to preserve the formatting}
  {Third author} {Affiliation3 \\ {\tt author3@ismir.edu}}
% what order do we use? Thierry, Ron, Dan?  Dan, Thierry, Ron?

@@

\subsection{Clustering}

% We can then stack beat vectors for a number of bars (usually 1 or 2).

We use an online version of the vector quantization algorithm
\cite{Gersho1991} to cluster the beat-chroma patches described in the
previous section.
% This can also be seen as an online K-means.
For each sample from the data, the algorithm finds the closest cluster
in the codebook and updates the cluster centroid (codeword) to be
closer to the sample, using a learning rate parameter $\ell$.
The clusters are updated as each data point is seen, as opposed to
once per iteration in the standard k-means algorithm.
The details can be found in Algorithm \ref{algo:vq}.
%
As in standard k-means clustering, the codebook is initialized by
choosing $K$ random points from our dataset.
%
Note that this algorithm, although not optimal, scales linearly with
the number of patches seen and can be interrupted at any time
while still resulting in an updated codebook. %not that this is important...
% The setting of the learning rate might become trickier, but this is
% not fundamentally a problem.

@@

\section{Experiments}\label{sec:experiments}

% In this section we present different clustering experiments. We introduce
% our training data and our main testing data. Some detailed settings
% of our algorithm are also provided. As for any clustering algorithm, we
% measure the influence of the number of codewords and the training set size.
In this section we present a set of experiments designed to evaluate
quantization quality as a function of size of training set, codebook
size, and patch length.

\subsection{Data}
%\subsection{Training data: Cowbell Dataset}
\label{sec:traindata}
%\subsection{Test data: USPOP}
\label{sec:testdata}
The training data consists of $43,300$ tracks that were uploaded to
morecowbell.dj,\footnote{http://www.morecowbell.dj/} an online service
based on the Echo Nest analyze API which remixes music by
adding cowbell and other sound effects synchronized in time with the
music.  The $43.3$K songs contain $\mathbf{3.7}$ million non silent
bars which we cluster using the approach described in the previous
section.

For testing, we made use of low quality (32kbps MP3) audio 
from the uspop2002 dataset \cite{uspop2002}.  The collections
includes a wide variety of popular music from 10 different sub-genres.
% \footnote{We also tried to use Tzanetakis dataset,
% %\cite{Tzanetakis2002a},
% but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
% most did not contain any bar.}.
We use this data to measure how well a codebook learned on
the Cowbell dataset can represent new songs.  
We obtained Echo Nest features for the $8651$ songs contained in the dataset.

\subsection{Setting}\label{ssec:setting}
We train codebooks based on beat-chroma patches composed of 4 or 8
beats (one or two bars).
Codebooks over the Cowbell
dataset are trained using the VQ algorithm (Algorithm \ref{algo:vq}). We used a
learning rate of $\ell=0.01$ for $200$ iterations over the data.
%
Given these codebooks, we quantize the test set by assigning each
patch to a single codeword.  We report the average
distance between patches and their encodings and describe the usage
frequency of the different codes. %, \ie the size of the cluster around a code.

We use the average squared Euclidean distance as the distortion
measure between chroma patches.  Given a pattern $p_1$
composed of elements $p_{1ij}$, and a second pattern $p_2$, the
distance between them is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{(p_{1ij} - p_{2ij})^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume that $p_1$ and $p_2$ have the same size.  This is enforced by the
meter normalization procedure described in Section \ref{sec:approach}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time, a bar being represented by $4$ chroma feature vector.
Therefore, a codeword or a pattern is of size $12$ x $4$ = $48$.
Distortion measure on the test dataset. Data size tested range
from $0$ (just initialization) to $500$ thousand. Patterns were selected at
random from the dataset of approximately $3,7$ million patterns.
}}
\label{fig:data_sizes}
\end{figure}

\subsection{Basic Properties}
This section present some fundamental results of the clustering algorithm.
None of the result appear surprising,  but they are important for 
reproducing the experiments.
\begin{itemize}
\item Encoding performance improves with increasing training data (Figure
\ref{fig:data_sizes}). Regarding the convergence of the performance,
we can imagine using $500$ or $1$K samples per codeword to learn if the
training set size is problematic.
\item Encoding performance improves with increasing codebook size
(Table \ref{tab:cbsize}). The increase seems significant, and
increasing the size of the codebook in an efficient way
is definitely part of our future work.
\item Larger patterns are more difficult to encode, thus requiring
larger codebooks. See Figure \ref{fig:perbeat}. The increase is steeper
until $4$ beats which usually corresponds to one bar.
\end{itemize}

\begin{table}
\begin{center}
\begin{tabular}{llc}
\toprule
Codebook size & Distortion \\
\midrule
1 & $0.066081$ \\
10 & $0.045579$ \\
50 & $0.038302$ \\
100 & $0.035904$ \\
500 & $0.030841$ \\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{Varying codebook size using $50,000$ training samples.
    We train on patterns corresponding to $1$ bar resized to $4$ beats.
  }}
\label{tab:cbsize}
\end{table}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding larger and larger patterns with a fixed size
codebook. The size of the pattern is defined by the number of beats,
the bar information was not taken into account for this experiment.
}}
\label{fig:perbeat}
\end{figure}

@@

We train a codebook of size $200$ over patterns of size $12$x$8$ covering
$2$ bars at a time.  The results shown are on the test set from
Section \ref{ssec:artist}.

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords are shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 0),
perfect fifth (codewords 1 and 2) and fourth intervals (codewords 3
and 6), and 
major and minor triads (codewords 5 and 4, respectively).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a tonic-subdominant transition in codes 7 and 9
(Cmaj
$\rightarrow$ Fmaj or C5 $\rightarrow$ F5 as a guitar power chord)
% assuming that chroma 0 is C) 
and the reverse tonic-dominant transition in code 21 (\ie Fmaj
$\rightarrow$ Cmaj).
%FIXME: Does this transition have a better name like ``relative x''?
% C5 -> F5 in 'guitar chords', basically a power chord

\begin{figure}[h]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes1}
\end{center}
\caption{\small{
25 most frequently used codes in the codebook %in term of frequency of use
in the artist20 test set. 
This codebook contains $200$ codes trained on $2$ bar, $12 \times 8$
patches.
The numbers 
%above each panel correspond to the display are the index (from most used to least) and
in parentheses correspond to 
the percentage of use in the test set.
%There are $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage frequency of the patterns in the codebook from
Figure \ref{fig:codes1} on the artist20 dataset.  There are $71,832$
patterns in total in the dataset. Also frequency on the training set.
Note that even though all codewords are initialized from samples, some
are used only once (training set) or never (test set).
}}
\label{fig:freqs}
\end{figure}

@@

Figure \ref{fig:encodesong} shows an example of a song encoded
using the codebook described in Section \ref{sec:codebook}.  Both the
Echo Nest chroma features for an example song and the corresponding
quantization are displayed.  The quantized representation retains
dominant harmonic structure of the original features, but is smoothed
in time due to the use of 2 bar long codewords.
%
This is consistent with the codebook described in Section
\ref{sec:codebook} which primarily consists of sustained chords across
4 or 8 beats.

>>>>

% Three addresses
% --------------
\threeauthors
  {First author} {Affiliation1 \\ {\tt author1@ismir.edu}}
  {Second author} {\bf Retain these fake authors in\\\bf submission to preserve the formatting}
  {Third author} {Affiliation3 \\ {\tt author3@ismir.edu}}
% Thierry, Ron, Dan

@@

\subsection{Clustering}
% We can then stack beat vectors for a number of bars (usually 1 or 2).

We use an online version of the vector quantization algorithm
\cite{Gersho1991} to cluster the beat-chroma patches described in the
previous section.
% This can also be seen as an online K-means.
For each sample from the data, the algorithm finds the closest cluster
in the codebook and updates the cluster centroid (codeword) to be
closer to the sample according to a learning rate $\ell$.
The clusters are updated as each data point is seen, as opposed to
once per iteration in the standard $k$-means algorithm.
The details are explained in Algorithm \ref{algo:vq}.
%
As in standard $k$-means clustering, the codebook is initialized by
choosing $K$ random points from our dataset.
%
Note that this algorithm, although not optimal, scales linearly with
the number of patches seen and can be interrupted at any time
to obtain an updated codebook. %not that this is important...
% The setting of the learning rate might become trickier, but this is
% not fundamentally a problem.

@@

\section{Experiments}\label{sec:experiments}
In this section we present different clustering experiments and introduce
our principal training and test data. Some detailed settings
of our algorithm are also provided. As for any clustering algorithm, we
measure the influence of the number of codewords and the training set size.

\subsection{Data}
%\subsection{Training data: Cowbell Dataset}
\label{sec:traindata}
%\subsection{Test data: USPOP}
\label{sec:testdata}
Our training data consists of $43,300$ tracks that were uploaded to
morecowbell.dj,\footnote{http://www.morecowbell.dj/} an online service
based on the Echo Nest analyze API which remixes uploaded music by
adding cowbell and other sound effects synchronized in time with the
music.  The $43.3$K songs contain $3.7$ million non-silent
bars which we clustered using the approach described in the previous
section.

For testing, we made use of low quality (32kbps, 8 kHz bandwidth mono MP3) versions of the songs
from the {\it uspop2002} data set \cite{uspop2002}.  This well-known data set 
contains pop songs from a range of artists and styles.
% \footnote{We also tried to use Tzanetakis dataset,
% %\cite{Tzanetakis2002a},
% but the $30$ seconds segments seemed hard to analyze by the Echo Nest API, 
% most did not contain any bar.}.
{\it uspop2002} serves as test set to measure how well a codebook learned on
the Cowbell data set can represent new songs.  We obtained Echo Nest features
for the $8651$ songs contained in the dataset.

\subsection{Setting}\label{ssec:setting}
We take one or two bars and resample the patches to 4 or 8 columns 
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the online VQ algorithm (Algorithm \ref{algo:vq}). We use a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We then use the resulting codebook to encode the test set.
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, 
%\ie the size of the cluster around a code.
i.e., the proportion of patterns that quantize to each code.

We use the average squared Euclidean distance as the distortion
measure between chroma patches.  Given a pattern $p_1$
composed of elements $p_1(i,j)$, and a similar pattern $p_2$, the
distance between them is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{(p_1(i,j) - p_2(i,j))^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume $p_1$ and $p_2$ have the same size.  This is enforced by the
resampling procedure described in Section \ref{sec:approach}.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time with by $4$ columns.
Therefore, each codeword has $12 \times 4 = 48$ elements.
Distortion is measurde on the test set.  Training data sizes range
from $0$ (just initialization) to $500,000$. Patterns were selected at
random from the dataset of approximately $3.7$ million patterns.
}}
\label{fig:data_sizes}
\end{figure}

\subsection{Codebook properties}
This section presents some basic results of the clustering.
While unsurprising, these results may be useful for comparison 
when reproducing this work.
\begin{itemize}
\item Encoding performance improves with increasing training data (Figure
\ref{fig:data_sizes}). Distortion improvements plateau by around $1000$ samples 
per codeword ($100,000$ samples for the $100$-entry codebook of the figure).
\item Encoding performance improves with increasing codebook size
(Table \ref{tab:cbsize}).  Computation costs scales with codebook size, 
which limited the largest codebooks used in this work, but larger codebooks 
(and more efficient algorithms to enable them) are clearly a promising future 
direction.
\item Larger patterns are more difficult to encode, thus requiring
larger codebooks. See Figure \ref{fig:perbeat}. The increase is steepest
below $4$ beats (1 bar), although there is no dramatic change at 
this threshold.
\end{itemize}

\begin{table}
\begin{center}
\begin{tabular}{llc}
\toprule
Codebook size & Distortion \\
\midrule
1 & $0.066081$ \\
10 & $0.045579$ \\
50 & $0.038302$ \\
100 & $0.035904$ \\
500 & $0.030841$ \\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{Distortion as a function of codebook size for a fixed training set 
of $50,000$ samples.  Codebook consists of $1$ bar ($4$ beat) patterns.
  }}
\label{tab:cbsize}
\end{table}

\begin{figure}[b]
\begin{center}
\includegraphics[width=.8\columnwidth]{encoding_per_beat}
\end{center}
\caption{\small{Encoding patterns of different sizes with a fixed size
codebook of 100 patterns. The size of the pattern is defined by the number of beats. 
Downbeat (bar alignment) information was not used for this experiment. % why not?
}}
\label{fig:perbeat}
\end{figure}

@@

We trained a codebook $200$ entries of patterns sized $12\times8$, covering
$2$ bars at a time.  The results shown are on the {\it artist20} test set described in 
Section \ref{ssec:artist}.

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords is shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 0),
perfect fifth (codewords 1 and 2) and fourth intervals (codewords 3
and 6, noting that the fourth occurs when the per-pattern transposition 
detects the fifth rather than the root as the strongest chroma bin, and vice-versa), and a
major triads transposed to the root and fifth (codewords 5 and 4, respectively).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a V-I transition in codes 7 and 9
(e.g., Gmaj
$\rightarrow$ Cmaj or G5 $\rightarrow$ C5 as a guitar power chord)
% assuming that chroma 0 is C) 
and the reverse I-V transition in code 21 (e.g., Cmaj
$\rightarrow$ Gmaj).
%FIXME: Does this transition have a better name like ``relative x''?
% C5 -> F5 in 'guitar chords', basically a power chord

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes1}
\end{center}
\caption{\small{
The 25 codes that are 
most commonly used for the {\it artist20} test set. 
Codes are from the $200$-entry codebook trained on $2$ bar, $12 \times 8$
patches.  The proportion of patches accounted for by each pattern is shown 
in parentheses.
%There are $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

\begin{figure}[bt]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage proportions for all 200 codewords 
on the {\it artist20} test set (which comprises $71,832$
patterns).  Also shown are the usage proportions for the training set (``cowbell''), 
which are similar.
Note that even though all codewords are initialized from samples, some
are used only once in the training set, or not at all for test set.
}}
\label{fig:freqs}
\end{figure}

@@

Figure \ref{fig:encodesong} shows an example of a song encoded
using the codebook described in Section \ref{sec:codebook}.  Both the
Echo Nest chroma features for the example song and the corresponding
quantization are displayed.  The quantized representation retains
dominant harmonic structure of the original features, but is smoothed
in time due to the use of 2 bar long codewords.
%
This is consistent with the codebook described in Section
\ref{sec:codebook} which primarily consists of sustained chords across
4 or 8 beats.
